### 2025-06-09 manuscript/main.tex
restructuring: model checking, prob.  Symbolic eval

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage{fontspec}

@@

\section{Language Structure and Paradigms}
\subsection{Abstract vs. Concrete Syntax}
Depednming on subsection Grammar Formalisms, mentino the importance of grammars in connectino with programming langs.

@@

In domains where fine-grained control and explicit ordering mat
for choosing between replacing or recalibrating components—but measured data often reflect intertwined effects and closed-loop corrections. reating a declarative one: typically, an interpreters process each command in sequence, 
or a translator converts DSL instructions into a host general-purpose language.

Mybe just leave the below out

Imperative DSLs shine in scripting and macro contexts, where users sequence domain-specific commands to drive behavior, examples include Excel macros that
automate spreadsheet tasks or scripts within a game engine that dictate character actions and game logic. They are also well suited to testing frameworks
and scenario specifications, where user interfaces describe interactions by specifying a timely sequence of user inputs. In robotics and industrial automation, 
imperative DSLs allow practitioners to program custom action sequences — such as moving a robotic arm
to pick up an object, then placing it precisely — where each step must follow a strict order to ensure safety and accuracy. More broadly, any process-control 
domain that relies on precise sequencing and stateful interactions — be it a complex simulation workflow, an interactive operation pipeline, or automated machinery
— benefits from the clarity and control that an impera
for choosing between replacing or recalibrating components—but measured data often reflect intertwined effects and closed-loop corrections. tive DSL provides.

\chapter{Probabilistic Programming \& Cyber-Physical Systems}
\section{What Is Probabilistic Programming?}

@@

\textit{Bayes’ Theorem} provides the foundational framework for probabilistic reasoning under uncertainty. It describes how to update prior beliefs about unknown 
quantities - known as latent variables - based on new evidence, resulting in a posterior distribution. Formally, it states that the posterior is proportional to the
product of the prior and the likelihood:
\[
P(\theta \mid x) \propto P(x \mid \theta) \cdot P(\theta)
\]
In probabilistic programming, this principle is applied through models that define prior distributions over latent variables and likelihood functions for 
observed data. The probabilistic programming language then applies Bayes' Theorem, typically via automated inference algorithms, to compute or approximate the posterior.
This makes it possible to express complex probabilistic models declaratively, while delegating the computation of posterior beliefs to the underlying inference engine.
\\
??INLUDE GRAPHICAL MODESL??

@@

Sampling-based methods approximate the posteriIn probabilistic programming, t

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{graphics/inferencePlot.png}
  \caption{Comparison of exact posterior to inference methods}
  \label{fig:inference_plot}
\end{figure}

\section{Motivation for PPLs in CPS Contexts}
??enumerate the four paragrpahs into 2x2??

When measurements contain noise or are incorrect, PPL models can still be applied to extract useful information. Instead of treating the raw reading of a sensor as the “truth,”
the readings are considered samples for a \textit{sensor’s model}. Not only is it possible to reduce the noise in the measurements by calculating the posterior, 
but those noise characteristics are also used to flesh out the predictive model itself.
This enables PPLs to adaptively adjust noise models online when, for example, sensors deteriorate over time.
Additionally — with most CPSs employing multiple sensors — a PPL naturally supports \textit{sensor fusion}, \cite{cpsSensorFusion} combining data from heterogeneous sources
to produce a coherent estimate. By defining each sensor's observation as a conditional distribution, a probabilistic program
weights each reading according to its uncertainty: more reliable sensors contribute more heavily, while inconsistent or faulty readings are down-weighted or flagged. 
Inference then yields a joint posterior over the true state, allowing the system to reconcile disagreements between sensors, detect potential sensor failures by inferring 
latent bias variables, and maintain a quantified confidence.\\

Which leads us into robotics, a quintessential cornerstone of CPS. Instead of working on raw observational data and making assumptions in incomplete cases, robots can adopt 
a \textit{belief} based on an observational model \cite{cpsProbabilisticRobotics}. The observational model couples motion dynamics (how actions translate into state 
transitions) with existing sensor models, providing a full posterior distribution that the robot can use to imply different levels of confidence for its actions.
Furthermore, robots often operate in changing environments or with components that degrade over time. Within a PPL, unknown parameters, such as friction 
coefficients or sensor calibration offsets, can be treated as latent variables and perform Bayesian learning to infer them from incoming data. The same inference mechanism used for state 
estimation thus also adapts the model online, continuously refining both motion dynamics and sensor mappings.\\

A PPL expresses system dynamics probabilistically so that sampling yields a distribution of possible futures. \cite{cpsPredictiveControl} By enforcing “chance constraints” 
(e.g., “limit the probability of exceeding a safety threshold to 1\%”), inference or Monte Carlo sampling identifies control inputs that optimize expected performance while 
respecting risk. When the current state is uncertain, planning occurs in belief space — each action triggers a Bayesian update of the state distribution, and planners 
choose sequences that minimize expected cost given that uncertainty. Simultaneously, online Bayesian learning refines model parameters (like friction or drag) over time, 
tightening future predictions; as uncertainty shrinks, control can push closer to limits, whereas rising uncertainty leads to more conservative actions.\\

Recent research is looking at ways to utilize PPLs to gather diagnostics of a CPS. \cite{cpsPPLDiagnostics} They authors propose a way to infer potential error causes, that
might be overlooked otherwise, due to the . 
This of of great interest to system operators for deciding if either parts have to be replaced or only have to be recelebrated. Because measured data on system deviations
often reflects complex intertwining causes and \textit{closed-loop corrections}, identifying root causes becomes a matter of inference. The proposed two-step approach 
first builds a generative model (based on control software and expert insights) that simulates observations from hypothesized error causes. This simulator is then recast 
as a probabilistic program, allowing Bayesian inference to estimate latent causes (with confidence intervals) from actual measurements.

\chapter{The \textit{Miniprob} DSL}
\section{Domain and Purpose}
\section{Syntax Overview}
\section{Semantics and Example Models}

>>>>

\documentclass[11pt]{report}
\usepackage[paper=a4paper,margin=2.5cm]{geometry}
\usepackage{multicol}
\usepackage{fontspec}

@@

\section{Language Structure and Paradigms}
\subsection{Abstract vs. Concrete Syntax}
Depednming on subsection Grammar Formalisms, mentino the importance of grammars in connectino with programming langs.??

@@

In domains where fine-grained control and explicit ordering matter, imperative DSLs provide maximal flexibility, because every aspect of execution is adjustable.
This level of control becomes crucial when timing or order of operations yield different outcomes, or when fine-tuned performance optimizations are required.
Furthermore, building an imperative DSL is often simpler than creating a declarative one: typically, an interpreters process each command in sequence, 
or a translator converts DSL instructions into a host general-purpose language. \cite{}

Imperative DSLs are well suited for scripting and defining macros, where users sequence domain-specific commands to drive behavior. They also shine in combination with testing
frameworks and scenario specifications, where user interfaces describe interactions by specifying a timely sequence of user inputs. In robotics and industrial automation, 
imperative DSLs allow to program custom action sequences — such picking up an object, then placing it precisely — where each step must follow a
strict order to ensure safety and accuracy. More broadly, any process-control domain that relies on precise sequencing and stateful interactions benefits from the clarity
and control that an imperative DSL provides.

\chapter{Cyber-Physical Systems and Formal Analysis}

\section{Formal Methods and Model Checking}

CPSs are often deployed in safety-critical environments where utmost importance is placed on reliable operation and functional correctness. Yet reaching high levels of confidence
in a systems behavior remains extremely inside cyber-physical domains.\cite{formalMethodsCPSCritical} This difficulty arises from the complex interactions of all the individual integral 
components of the system, which can not be effectively tested only employing traditional methods. The inherent entanglement of these systems warrants the use of
specialized procedures that are able to cover and validate the acting components. Unlike conventional simulation based testing, \textit{formal methods} are able to prove correctness against
formal specifications, thereby providing more reliable results and eliminating potential error sources, such as race conditions or violations of safety properties(unnoticed states).

In formal methods, \textit{theorem proving}, \textit{model checking}, and \textit{runtime verification} are widely recognized as the three main approaches for assuring system correctness. \cite{formalMethodsBig3,formalMethodsCPSCritical}
All three practices are distinguishable from one another and are tailored for certain scenarios. Runtime verification is a lightweight technique that verifies systems 
by inspecting their execution trace - often employing automata to detect violations. Theorem proving, in contrast, is an offline and often time intensive 
process relying on mathematical inference rules to statically validate correctness with a high degree of rigor.

Model checking is of most interest to us, as the MiniProb DSL \ref{sec:miniprob} was developed with this context in mind. \cite{THEONEANDONLY1}
Logical Formulas temporal formulas ...
States blabla ... ..

Model checking is an automated formal verification technique used to exhaustively analyze the behavior of finite-state systems against a formally specified set of 
properties. These properties are typically expressed in temporal logics such as Linear Temporal Logic (LTL) or Computation Tree Logic (CTL), which enable reasoning 
about sequences of states over time (Baier & Katoen, 2008). A model checker systematically explores all reachable states of the system model to verify whether the 
temporal logic specification holds. If a property is violated, the tool provides a counterexample trace, aiding in pinpointing the source of the issue (
  Clarke et al., 1999). The use of logical specifications allows model checking to offer strong correctness guarantees, particularly for safety and liveness 
  properties in reactive and concurrent systems.

\cite{formalMethodsCPSCritical} explains model checking

\section{Motivation for PPLs in CPS Contexts}

\begin{multicols}{2}
When measurements contain noise or are incorrect, PPL models can still be applied to extract useful information. Instead of treating the raw reading of a sensor as the “truth,”
the readings are considered samples for a \textit{sensor’s model}. Not only is it possible to reduce the noise in the measurements by calculating the posterior, 
but those noise characteristics are also used to flesh out the predictive model itself.
This enables PPLs to adaptively adjust noise models online when, for example, sensors deteriorate over time.
Additionally — with most CPSs employing multiple sensors — a PPL naturally supports \textit{sensor fusion}, \cite{cpsSensorFusion} combining data from heterogeneous sources
to produce a coherent estimate. By defining each sensor's observation as a conditional distribution, a probabilistic program
weights each reading according to its uncertainty: more reliable sensors contribute more heavily, while inconsistent or faulty readings are down-weighted or flagged. 
Inference then yields a joint posterior over the true state, allowing the system to reconcile disagreements between sensors, detect potential sensor failures by inferring 
latent bias variables, and maintain a quantified confidence.
\\
Which leads us into robotics, a quintessential cornerstone of CPS. Instead of working on raw observational data and making assumptions in incomplete cases, robots can adopt 
a \textit{belief} based on an observational model \cite{cpsProbabilisticRobotics}. The observational model couples motion dynamics (how actions translate into state 
transitions) with existing sensor models, providing a full posterior distribution that the robot can use to imply different levels of confidence for its actions.
Furthermore, robots often operate in changing environments or with components that degrade over time. Within a PPL, unknown parameters, such as friction 
coefficients or sensor calibration offsets, can be treated as latent variables and perform Bayesian learning to infer them from incoming data. The same inference mechanism used for state 
estimation thus also adapts the model online, continuously refining both motion dynamics and sensor mappings.
\\
A PPL expresses system dynamics probabilistically so that sampling yields a distribution of possible futures. \cite{cpsPredictiveControl} By enforcing “chance constraints” 
(e.g., “limit the probability of exceeding a safety threshold to 1\%”), inference or Monte Carlo sampling identifies control inputs that optimize expected performance while 
respecting risk. When the current state is uncertain, planning occurs in belief space — each action triggers a Bayesian update of the state distribution, and planners 
choose sequences that minimize expected cost given that uncertainty. Simultaneously, online Bayesian learning refines model parameters (like friction or drag) over time, 
tightening future predictions; as uncertainty shrinks, control can push closer to limits, whereas rising uncertainty leads to more conservative actions.
\\
Recent research is looking at ways to utilize PPLs to gather diagnostics of a CPS. \cite{cpsPPLDiagnostics} They authors propose a way to infer potential error causes, that
might be overlooked otherwise, due to the . 
This of of great interest to system operators for deciding if either parts have to be replaced or only have to be recelebrated. Because measured data on system deviations
often reflects complex intertwining causes and \textit{closed-loop corrections}, identifying root causes becomes a matter of inference. The proposed two-step approach 
first builds a generative model (based on control software and expert insights) that simulates observations from hypothesized error causes. This simulator is then recast 
as a probabilistic program, allowing Bayesian inference to estimate latent causes (with confidence intervals) from actual measurements.

\end{multicols}

\chapter{Probabilistic Programming}
\section{What Is Probabilistic Programming?}

@@

\textit{Bayes’ Theorem} provides the foundational framework for probabilistic reasoning under uncertainty. It describes how to update prior beliefs about unknown 
quantities - known as latent variables - based on new evidence, resulting in a posterior distribution. Formally, it states that the posterior is proportional to the
product of the prior and the likelihood:
\[
P(\theta \mid x) \propto P(x \mid \theta) \cdot P(\theta)
\]
In probabilistic programming, this principle is applied through models that define prior distributions over latent variables and likelihood functions for 
observed data. The probabilistic programming language then applies Bayes' Theorem, typically via automated inference algorithms, to compute or approximate the posterior.
This makes it possible to express complex probabilistic models declaratively, while delegating the computation of posterior beliefs to the underlying inference engine.
\\

@@

Sampling-based inference approximates the posterior by drawing samples from it, typically using Markov Chain Monte Carlo (MCMC) methods.
Continuously, random samples are generated from a given distribution - usually the prior or a conditional- and used to progressively approximate
the posterior with the model being executed repeatedly.
These approaches are flexible and can handle complex models, but they are often computationally expensive and require many samples to ensure accuracy.
The resulting samples can then be used to estimate expectations, marginal distributions, or event probabilities.

\subsection*{3. Variational Inference}

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.7\textwidth]{graphics/inferencePlot.png}
  \caption{Comparison sample of exact posterior to inference methods}
  \label{fig:inference_plot}
\end{figure}

\section{Symbolic Semantics and Model Checking}
pOPA ... ref to CPS Model Checking or put chapter after CPS

Inlcude graphic displaying sampling approximations compared to a symbolic state evaluation

\label{sec:miniprob}
\chapter{The \textit{MiniProb} DSL}
\section{Domain and Purpose}
\section{Syntax Overview}
\section{Semantics and Example Models}
